---
title: "NLSY97"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    fig.align: "center"
---

```{r}
#| label: setup
library(causalDisco)
```

We will analyze this dataset

```{r}
data(nlsdata)
summary(nlsdata)
```

A description of the dataset is given below

Table: Survey variables overview

| Variable name | Tier | Type    | Description |
|---------------|------|---------|-------------|
| r1_hhchildren | R1   | Numeric | Number of household members younger than 18 |
| r1_urban      | R1   | Binary  | Urban status (Urban / Non-urban) |
| r1_mcollege   | R1   | Binary  | Did biological mother attend college? (Yes / No) |
| r6_exercise   | R6   | Numeric | Days with 30+ minutes exercise / week |
| r6_depressed  | R6   | Binary  | Having felt depressed during the last month (Sometimes / No) |
| r6_docvisits  | R6   | Numeric | Number of doctor’s visits during the last year |
| r12_health    | R12  | Binary  | Perception of own health (Good / Poor) |
| r12_depressed | R12  | Binary  | Having felt depressed during the last month (Sometimes / No) |
| r12_docvisits | R12  | Numeric | Number of doctor’s visits during the last year |

Convert the two columns which are integer to numeric (should be done in the dataset...)

```{r}
nlsdata$r1_hhchildren <- as.numeric(nlsdata$r1_hhchildren)
nlsdata$r6_exercise <- as.numeric(nlsdata$r6_exercise)
```

We are mainly interested in understanding what affects whether a person is depressed.

We start by using the PC algorithm with the mutual information (conditional Gaussian) test from bnlearn

```{r}
pc_bnlearn <- pc(engine = "bnlearn", test = "mi_cg")
res_pc_bnlearn <- disco(data = nlsdata, method = pc_bnlearn)
plot(res_pc_bnlearn, layout = "kamada-kawai")
```

Similar, we could try the same for Tetrad.

```{r}
pc_tetrad <- pc(engine = "tetrad", test = "conditional_gaussian")
res_pc_tetrad <- disco(data = nlsdata, method = pc_tetrad)
plot(res_pc_tetrad, layout = "kamada-kawai")
```

We see Tetrad has much fewer orientations, which is what one usually sees when compared to bnlearn
(it orients too much?).

If we don't believe in the conditional gaussian assumption, we could use more general notions. Here we use
the Generalized Independence Noise test.

```{r}
pc_tetrad <- pc(engine = "tetrad", test = "gin")
res_pc_tetrad <- disco(data = nlsdata, method = pc_tetrad)
plot(res_pc_tetrad, layout = "kamada-kawai")
```

use BOSS instead/also?


# Temporal Knowledge

We by definition have a temporal ordering of the variables, which we encode below

```{r}
kn <- knowledge(
  nlsdata,
  tier(
    r1 ~ starts_with("r1_"),
    r6 ~ starts_with("r6_"),
    r12 ~ starts_with("r12_")
  )
)
plot(kn)
```

Now we can use this knowledge in the algorithms from above

```{r}
pc_bnlearn <- pc(engine = "bnlearn", test = "mi_cg")
res_pc_bnlearn <- disco(data = nlsdata, method = pc_bnlearn, knowledge = kn)
plot(res_pc_bnlearn)
```

Similar, we could try the same for Tetrad.

```{r}
pc_tetrad <- pc(engine = "tetrad", test = "conditional_gaussian")
res_pc_tetrad <- disco(data = nlsdata, method = pc_tetrad, knowledge = kn)
plot(res_pc_tetrad)
```


If we don't believe in the conditional gaussian assumption, we could use more general notions. Here we use
the Generalized Independence Noise test.

```{r}
pc_tetrad <- pc(engine = "tetrad", test = "gin")
res_pc_tetrad <- disco(data = nlsdata, method = pc_tetrad, knowledge = kn)
plot(res_pc_tetrad)
```
