---
title: "NLSY97"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    fig.align: "center"
---

```{r}
#| label: setup
library(causalDisco)
```

We will analyze this dataset

```{r}
data(nlsdata)
summary(nlsdata)
```

A description of the dataset is given below

Table: Survey variables overview

| Variable name | Tier | Type    | Description |
|---------------|------|---------|-------------|
| r1_hhchildren | R1   | Numeric | Number of household members younger than 18 |
| r1_urban      | R1   | Binary  | Urban status (Urban / Non-urban) |
| r1_mcollege   | R1   | Binary  | Did biological mother attend college? (Yes / No) |
| r6_exercise   | R6   | Numeric | Days with 30+ minutes exercise / week |
| r6_depressed  | R6   | Binary  | Having felt depressed during the last month (Sometimes / No) |
| r6_docvisits  | R6   | Numeric | Number of doctor’s visits during the last year |
| r12_health    | R12  | Binary  | Perception of own health (Good / Poor) |
| r12_depressed | R12  | Binary  | Having felt depressed during the last month (Sometimes / No) |
| r12_docvisits | R12  | Numeric | Number of doctor’s visits during the last year |

Convert the two columns which are integer to numeric (should be done in the dataset...)

```{r}
nlsdata$r1_hhchildren <- as.numeric(nlsdata$r1_hhchildren)
nlsdata$r6_exercise <- as.numeric(nlsdata$r6_exercise)
```

We are mainly interested in understanding what affects whether a person is depressed.

We start by using the PC algorithm with the mutual information (conditional Gaussian) test from bnlearn

```{r}
pc_bnlearn <- pc(engine = "bnlearn", test = "mi_cg")
res_pc_bnlearn <- disco(data = nlsdata, method = pc_bnlearn)
plot(res_pc_bnlearn, layout = "kamada-kawai")
```

Similar, we could try the same for Tetrad

```{r}
pc_tetrad <- pc(engine = "tetrad", test = "conditional_gaussian")
res_pc_tetrad <- disco(data = nlsdata, method = pc_tetrad)
plot(res_pc_tetrad, layout = "kamada-kawai")
```



